A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-24 linear.tex
minor edits
minor changes
general edits

\section{Quasi-global linear problem}\label{linear_problem}
We consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta f (r, z) = \delta f_1(r,z)\exp{(\ii k r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k$ is a real wavenumber such that $|kr|\gg 1$, and the
amplitude $\dd f_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

@@

\begin{align}
  \ii\sigma W &= \ii k \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k Q,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right]  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy} 
\end{align}  
where $^\prime \equiv \p_z$ and $\bm{F} \equiv -\nabla P/\rho$. 
The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
Note that we have assumed a temperature profile that only depends on $r$.

>>>>

\section{Linearized equations}\label{linear_problem}
To compute explicit solutions to the linear problem, 
we consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta f (r, z) = \delta f_1(r,z)\exp{(\ii k r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k$ is a real wavenumber such that $|kr|\gg 1$, and the
amplitude $\dd f_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

@@

\begin{align}
  \ii\sigma W &= \ii k \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k Q,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right]  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy} 
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla P/\rho$. 
The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
Note that we have assumed a temperature profile that only depends on $r$.

### 2016-08-24 model.tex
general edits

\subsection{Strictly isothermal gas perfectly coupled to dust}   
When $c_s^2$ is a constant and $\tstop=0$, the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit adiabatic
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\tepsilon$ is advected with 
the gas. This is similar to entropy being conserved following an
adiabatic gas when there is no heating or cooling.

@@

The Lagragian perturbation $\Delta$ of a variable $f$ is 
related to its Eulerian perturbation by $\Delta f = \delta f + 
\bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the Lagragian  
displacement ($\dd v_{x,z} = -\ii\sigma\xi_{x,z}$). 
%Note that $\Delta
%\mathcal{C} = \delta \mathcal{C}$ for a uniform background $C$ as
%required for equilibrium. 
Then Eq. \ref{lin_mass_full}---\ref{lin_energy_full} imply  
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). 
\end{align}
This reflects `$TdS  =  PdV$' from thermodynamics, if the 
internal energy is constant. That is, dust-gas drag provides 
heating ($TdS$) to do $PdV$ work. Whether positive or negative work is 
done depends on the correlation between the pressure and density
perturbations, as follows.

@@

We suggest understanding dusty instabilities in
protoplanetary disks by re-casting the problem into that of
thermodynamics. Then dust-drag instabilities appear in the equations in the 
same way as pulsational instabities in stars, which has already been 
well studied. Whether or not the dusty disk is actually unstable
depends on details of the dust-drag or `cooling' function
$\mathcal{C}$. However, we might generally expect the system to
support a range of modes. On physical grounds, unstable modes 
\emph{must} be so by causing a phase lag between pressure and total
density.

@@

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
If $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel} 
gives  
\begin{align}
  s = \frac{\imag\int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check}
\end{align} 
This is classic VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We compute
explicit solutions in this limit in \S\ref{results}.

>>>>

\subsection{Strictly isothermal gas perfectly coupled to dust}\label{iso_perfect}
When $c_s^2$ is a constant and $\tstop=0$, the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit adiabatic
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\tepsilon$ is advected with 
the gas. This is similar to entropy being conserved following an
adiabatic gas when there is no heating or cooling.

@@

The Lagragian perturbation $\Delta$ of a variable $f$ is 
related to its Eulerian perturbation by $\Delta f = \delta f + 
\bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the Lagragian  
displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$). 
%Note that $\Delta
%\mathcal{C} = \delta \mathcal{C}$ for a uniform background $C$ as
%required for equilibrium. 
Then Eq. \ref{lin_mass_full}---\ref{lin_energy_full} imply  
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). 
\end{align}
This reflects `$TdS  =  PdV$' from thermodynamics, if the 
internal energy is constant. That is, dust-gas drag provides 
heating ($TdS$) to do $PdV$ work. Whether positive or negative work is 
done depends on the correlation between the pressure and density
perturbations, as follows.

@@

We suggest understanding dusty instabilities in
protoplanetary disks by re-casting the problem into that of
thermodynamics. Then dust-drag instabilities appear in the equations in the 
same way as pulsational instabities in stars, which has already been 
well studied. Whether or not the dusty disk is actually unstable
depends on details of the dust-drag or `cooling' function
$\mathcal{C}$. However, given the complexity of $\mathcal{C}$ (
Appendix \ref{lin_dust}), we might generally expect the system to
support a range of modes, stable and unstable. 
On physical grounds, growing modes  
\emph{must} be so by causing a phase lag between pressure and total
density.

@@

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
If $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel} 
gives  
\begin{align}
  s = \frac{\imag\int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check}
\end{align} 
This is VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We present 
numerical solutions of the VSI with perfectly-coupled dust in \S\ref{results}.

### 2016-08-24 paper.tex
general edits

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This property is similar to 
  entropy  conservation
  following an adiabatic fluid.  We show that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty accretion disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show that in the limit of perfect coupling, dust settling in
itself cannot lead to axisymmetric instabilities, however thin the dust layer becomes.     
% 
We also extend 
the vertical shear instability (VSI) to locally isothermal, dusty 
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilize modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We discuss thermodynamical interpretations of 
dust-gas drag instabilities, for example the streaming
instability in protoplanetary disks, as a phase lag between the pressure 
and density of a dusty gas, which is   
analogous to pulsational instabilities 
in stars.  
\end{abstract}

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This property is analogous to 
  entropy  conservation
  following an adiabatic fluid.  We show that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty accretion disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show that in the limit of perfect coupling, dust settling in
itself cannot lead to axisymmetric instabilities, however thin the dust layer becomes.     
% 
We also extend 
the vertical shear instability (VSI) to locally isothermal, dusty 
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilize modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We discuss thermodynamical interpretations of 
dust-gas drag instabilities, for example the streaming
instability in protoplanetary disks, as a phase lag between the pressure 
and density of a dusty gas, much like pulsational instabilities 
in stars.  
\end{abstract}

### 2016-08-24 result.tex
minor edits
general edits

We calculate explicit solutions to the linearized
equations in the limit of perfectly coupled dust ($\tstop=0$), so that
the equilibria defined in \S\ref{eqm} are exact steady states. 
We use this to
examine the effect of dust on classic VSI, defined to be driven by the
temperature gradient (\S\ref{dusty_vsi_int}). For simplicity we
consider constant input values of the midplane dust-to-gas ratio
$\epsilon_0$ and the characteristic dust thickness $\Hd$. These,
together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used in \citetalias{lin15}.

Two boundary conditions are needed for the $\tstop=0$
problem. This is expected since this problem is exactly equivalent to
adiabatic hydrodynamics \citep[e.g.][]{lubow93}. For simplicity we
impose solid boundaries so that $\delta v_z(\pm\zmax)=0$.  
We solve the linearized equations as a generalized eigenvalue problem 
using a pseudo-spectral code adapted from \cite{lin15}. Amplitudes 
are expanded in Chebyshev polynomials up to order $N_z=512$. We check
results using Eq. \ref{vsi_check}.

\subsection{Qualitative expectations}
\citetalias{lin15} find the appropriate way to compare 
destabilizing vertical shear to stabilizing vertical buoyancy is
$r\p_z\Omega/\OmK$ against $N_z^2/\OmK^2$. We see from \S\ref{vertshear}
and \S\ref{vbuoyancy} that $|r\p_z\Omega|\sim q h_\mathrm{g}\OmK$ for a
thin, gas dominated disk; while $N_z^2\sim \epsilon\OmK^2$. Thus we
expect dust-induced buoyancy forces to stabilize the disk against the
VSI where $\epsilon \gtrsim h_\mathrm{g}$.

@@

We compare in Fig. \ref{compare_vshear_fixHd} the basic state
vertical shear rate, which is destabilizing, and the vertical buoyancy
frequency, which is stabilizing. For the nearly dust-free disk
$\epsilon_0=10^{-3}$ the vertical shear dominates buoyancy for all
$|z|>0$. However, a heavy dust-load with $\epsilon_0=1$ renders the
buoyancy to dominate over vertical shear in the disk atmosphere
$|z|\gtrsim 2.5$. We thus expect instability all heights for
$\epsilon_0=10^{-3}$, but to be restricted to the midplane for
$\epsilon_0=1$.

@@

Fig. \ref{vsi_dust_loading} show unstable modes for different values
of $\epsilon_0$ with fixed perturbation wavenumber  $k\Hg = 30$. The
eigenvalue distributions for $\epsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}, consisting
of the roughly horizontal `body modes' and the nearly-vertical
`surface modes'. The latter is due to the imposed vertical boundaries
\citep{barker15}.

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably `surface modes', which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$. The body modes' growth rates remain $\sim O(h_\mathrm{g}\OmK)$
but their oscillation frequency increases with
$\epsilon_0$. The total number of modes do not change
significantly. This is in contrast with the effect of increasing
cooling times, where \citetalias{lin15} find fewer unstable modes.
%until the system is eventually completely stable.

@@

The lowest frequency `fundamental' body mode is energetically dominant
because the entire disk column is perturbed \citep[cf. surface modes
  which only disturb the disk boundaries,][]{umurhan16c}. In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$, where most of the disk mass
reside. This is consistent with the above comparison of the basic state vertical
shear and buoyancy. Notice also the perturbed dust-to-gas ratio,
$\delta\epsilon$, shifts from being maximized at the disk boundaries
to the midplane.

@@

In Fig. \ref{vsi_dust_loading_vareps} we plot the growth rates as a
function of $\epsilon_0$ for different perturbation wavenumbers $k$. We
find that dust-loading stabilizes the VSI more effectively for shorter
wavelength perturbations. This is because for high wavenumbers the
dominant modes are surface modes, which are stabilized by
dust-loading. The figure suggest that VSI becomes much less efficient
for $\epsilon_0\gtrsim 0.1$ and $k\Hg\gtrsim 50$.

@@

\subsection{Effect of dust thickness} 
We now vary the dust layer thickness $\Hd$ but fix the metalicity 
$Z \equiv \epsilon_0 \Hd/\Hg = \text{constant}$. This then gives the
required midplane dust-to-gas ratio. Since we will consider thin dust
layers, here we use a smaller 
domain with $\zmax=2\Hg$ so that $\epsilon$ does not become
too small.

We set $Z=0.03$ and analyse two disks with $\Hd=0.1\Hg$ and 
$\Hd=0.99\Hg$. Fig. \ref{compare_vshear_fixZ} compares the vertical
vhear rate and buoyancy frequency. For $|z|\gtrsim 0.4\Hg$ the two
disks have the same profile with vertical shear dominating over
buoyancy. We thus expect instabilities away from the disk midplane in 
both cases. For $|z|\lesssim 0.4\Hg$, a thin dust 
layer with $\Hd=0.1\Hg$ boosts the vertical shear rate, but the
associated buoyancy is larger still. Thus we expect the mid-plane of
the disk with $\Hd=0.1\Hg$ to be stable.

@@

Fig. \ref{result2d_fixZ} compares the fastest-growing VSI body modes  
with $k\Hg=30$ for the two cases above.\citepalias[The thinner domain
  adopted here eliminates surface modes, ][]{lin15}.   
We find very similar mode 
frequencies 
\begin{align*}
  \sigma = \begin{cases}
    \left(0.3053\ii - 0.8142\right)h_\mathrm{g}\OmK & \Hd=0.99\Hg, \\
    \left(0.3178\ii - 1.2237\right)h_\mathrm{g}\OmK & \Hd=0.1\Hg,
  \end{cases}
\end{align*}
since the vertical shear profile is similar throughout most of the
disk. However, meridional motions are surpressed near the midplane of
the $\Hd=0.1\Hg$ disk, as expected from the larger buoyancy frequency
as compared to vertical shear there. This leads to a structure
analogous to PPD dead zones: a quiescent midplane between active
surface layers \citep{gammie96}.

@@

Fig. \ref{compare_eigenvals_fixZ} shows the maximum VSI growth rates
as a function of the dust layer thickness. As before, we find growth
rates are most affected by the vertical structure of the dust layer
when the perturbation wavenumer is large. VSI growth rates converge as
$\Hd\to 0$. %to dust free values?    
Thus a thin dust layer, however large its associated vertical shear,
does not affect the VSI. The non-monotonic behavior for $\Hd\gtrsim
0.5\Hg$ arises because the vertical buoyancy frequency 
\begin{align*}
N_z^2(H_d;z,Z) \simeq &Z\Hg z^2
\exp{\left(-\frac{z^2}{2\Hg^2}\right)}\OmK^2\notag\\
&\times 
\frac{1}{\Hd}\left(\frac{1}{\Hd^2} -
\frac{1}{\Hg^2}\right)\exp{\left(-\frac{z^2}{2\Hd^2}\right)}  
\end{align*}
is a non-monotonic function of $\Hd$ at fixed $z$. At $z=\Hg$ and $z=2\Hg$
the buoyancy frequency is maximized for $\Hd\simeq0.5\Hg$ and
$\Hd\simeq 0.8\Hg$, respectively. This is consistent with the abscissa
of minima in growth rates in Fig. \ref{compare_eigenvals_fixZ}.

@@

\subsection{Validity}
We have considered $\tstop=0$ for simplicity. When $\tstop\neq0$, a 
disk initialized as in \S\ref{eqm} will evolve 
due to dust-gas relative drift. In particular, in the
absence of turbulent/diffusive processes, dust particles settle to the
midplane on a timescale $t_\mathrm{settle}\OmK \sim 1/\tstop\OmK$
\citep{takeuchi02}. For the previous results to be valid (i.e. the
basic state does not evolve significantly) we require VSI growth
timescales $t_\mathrm{grow}\ll t_\mathrm{settle}$. Since the maximum
VSI growth rate is $\sim h_\mathrm{g}\OmK$, we need stopping times
such that 
\begin{align}
  \tstop\OmK \ll h_\mathrm{g}. 
\end{align} 
For thin PPD with $h_\mathrm{g}\sim 0.05$, our dusty VSI results are
valid for $\tstop\OmK < O(10^{-2})$.

>>>>

We present numerical solutions to the linearized
equations to examine the effect of dust on the VSI. 
We consider perfectly coupled dust ($\tstop=0$) so that the 
equilibria defined in \S\ref{eqm} are exact steady states.

For simplicity we input constant values of the midplane dust-to-gas ratio
$\epsilon_0$ and the characteristic dust thickness $\Hd$. These,
together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used by 
\citet[][\citetalias{lin15} in this section]{lin15}.

Two boundary conditions are needed for the $\tstop=0$
problem. This is expected since this problem is equivalent to
adiabatic hydrodynamics \citep[e.g.][]{lubow93}. We 
impose solid boundaries so that $\delta v_z(\pm\zmax)=0$.  
We solve the linearized equations as a generalized eigenvalue problem 
using a pseudo-spectral code adapted from \cite{lin15}. Amplitudes 
are expanded in Chebyshev polynomials up to order $N_z=512$. We check
results using Eq. \ref{vsi_check}.

\subsection{Qualitative expectations}\label{vsi_est}
\cite{lin15} found the appropriate way to compare 
destabilizing vertical shear to stabilizing vertical buoyancy is
$r\p_z\Omega/\OmK$ against $N_z^2/\OmK^2$. We see from \S\ref{vertshear}
and \S\ref{vbuoyancy} that $|r\p_z\Omega|\sim q h_\mathrm{g}\OmK$ for a
thin, gas dominated disk; while $N_z^2\sim \epsilon\OmK^2$. Thus we
expect dust-induced buoyancy forces to stabilize the disk against the
VSI where $\epsilon \gtrsim h_\mathrm{g}$.

@@

Fig. \ref{compare_vshear_fixHd} compares the basic state
vertical shear rate, which is destabilizing, and the vertical buoyancy
frequency, which is stabilizing. For the nearly dust-free disk
$\epsilon_0=10^{-3}$ the vertical shear dominates buoyancy for all
$|z|>0$. However, a heavy dust-load with $\epsilon_0=1$ renders the 
buoyancy to dominate over vertical shear in the disk atmosphere 
$|z|\gtrsim 2.5$. We thus expect instability all heights for 
$\epsilon_0=10^{-3}$, but to be restricted to the midplane for
$\epsilon_0=1$.

@@

Fig. \ref{vsi_dust_loading} show unstable modes for different values
of $\epsilon_0$ with fixed perturbation wavenumber  $k\Hg = 30$. The
eigenvalue distributions for $\epsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}. This is
expected since $\epsilon_0 < \hgas$ (\S\ref{vsi_est}). Eigenvalues 
consists 
of the roughly horizontal `body modes' and the nearly-vertical
`surface modes'. The latter is due to the imposed vertical boundaries
\citep{barker15}.

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably, `surface modes', which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$ (i.e. $\epsilon_0> \hgas$).  
The body modes' growth rates remain $ O(h_\mathrm{g}\OmK)$ 
but their oscillation frequency increases with
$\epsilon_0$. %i.e. increases with buoyancing freq 
The total number of modes do not change
significantly. This is in contrast with the effect of increasing
cooling times, where \citetalias{lin15} find fewer unstable modes.
%until the system is eventually completely stable.

@@

The lowest frequency `fundamental' body mode is energetically dominant
because the entire disk column is perturbed \citep[cf. surface modes
  which only disturb the disk boundaries,][]{umurhan16c}. In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$. 
This is consistent with Fig. \ref{compare_vshear_fixHd} comparing the basic state vertical
shear and buoyancy. Notice also the perturbed dust-to-gas ratio,
$\delta\epsilon$, shifts from being maximized at the disk boundaries
to the midplane.

@@

In Fig. \ref{vsi_dust_loading_vareps} we plot the growth rates as a
function of $\epsilon_0$ for different perturbation wavenumbers $k$. We
find that dust-loading stabilizes the VSI more effectively for shorter
wavelength perturbations. This is because for high wavenumbers the
dominant modes are surface modes, which are effectively stabilized by
dust-loading since bouyancy forces are larges near the disk
boundaries. The figure suggest that VSI becomes much less efficient
for $\epsilon_0\gtrsim 0.1$ and $k\Hg\gtrsim 50$.

@@

\subsection{Effect of dust thickness} 
We now vary the dust layer thickness $\Hd$ but fix the metalicity 
$Z \equiv \epsilon_0 \Hd/\Hg = 0.03$. This then gives the
required midplane dust-to-gas ratio. Since we will consider thin dust
layers, here we use a smaller 
domain with $\zmax=2\Hg$ so that $\epsilon$ does not become
too small.

We analyze two disks with $\Hd=0.1\Hg$ and 
$\Hd=0.99\Hg$. Fig. \ref{compare_vshear_fixZ} compares the vertical
vhear rate and buoyancy frequency. For $|z|\gtrsim 0.4\Hg$ the two
disks have the same profile with vertical shear dominating over
buoyancy. We thus expect perturbations away from the disk midplane in 
both cases. For $|z|\lesssim 0.4\Hg$, a thin dust 
layer with $\Hd=0.1\Hg$ boosts the vertical shear rate, but the
associated buoyancy is larger still, implying the mid-plane should be
stable. 
%Thus we expect the mid-plane of 
%the disk with $\Hd=0.1\Hg$ to have limited vertical motions.

@@

Fig. \ref{result2d_fixZ} compares the fastest growing VSI body modes  
with $k\Hg=30$ for the two cases above.\citepalias[The thinner domain
  adopted here eliminates surface modes, ][]{lin15}.   
We find very similar mode 
frequencies 
\begin{align*}
  \sigma = \begin{cases}
    \left(0.3053\ii - 0.8142\right)h_\mathrm{g}\OmK & \Hd=0.99\Hg, \\
    \left(0.3178\ii - 1.2237\right)h_\mathrm{g}\OmK & \Hd=0.1\Hg,
  \end{cases}
\end{align*}
since the vertical shear profile is similar throughout most of the
disk. However, meridional motions are surpressed near the midplane of
the $\Hd=0.1\Hg$ disk, as expected from the larger buoyancy frequency
as compared to vertical shear there. This leads to a structure
analogous to PPD dead zones: a quiescent midplane between active
surface layers \citep{gammie96}.

@@

Fig. \ref{compare_eigenvals_fixZ} shows the maximum VSI growth rates
as a function of $\Hd$. As before, we find growth
rates are most affected by the vertical structure of the dust layer
when the perturbation wavenumer is large. Notice VSI growth rates converge as
$\Hd\to 0$. %to dust free values?    
Thus a thin dust layer, however large its associated vertical shear,
does not affect the VSI. The non-monotonic behavior for $\Hd\gtrsim
0.5\Hg$ arises because the vertical buoyancy frequency 
\begin{align*}
N_z^2(H_d;z,Z) \simeq &Z\Hg z^2
\exp{\left(-\frac{z^2}{2\Hg^2}\right)}\OmK^2\notag\\
&\times 
\frac{1}{\Hd}\left(\frac{1}{\Hd^2} -
\frac{1}{\Hg^2}\right)\exp{\left(-\frac{z^2}{2\Hd^2}\right)}  
\end{align*}
is a non-monotonic function of $\Hd$ at fixed $z$. At $z=\Hg$ and $z=2\Hg$
the buoyancy frequency is maximized for $\Hd\simeq0.5\Hg$ and
$\Hd\simeq 0.8\Hg$, respectively. This is consistent with the abscissa
of minima in growth rates in Fig. \ref{compare_eigenvals_fixZ}.

@@

\subsection{Validity}
We considered $\tstop=0$ in order to perform stability analysis on an
exact steady state. When $\tstop\neq0$, a 
disk initialized as in \S\ref{eqm} will evolve 
due to dust-gas relative drift. In particular, in the
absence of turbulent/diffusive processes, dust particles settle to the
midplane on a timescale $t_\mathrm{settle}\sim 1/\tstop\OmK^2$
\citep{takeuchi02}. However, we expect the above results to remain
valid provided the VSI growth
timescale $t_\mathrm{grow}\ll t_\mathrm{settle}$. Since 
VSI growth rates are $O(h_\mathrm{g}\OmK)$, we require 
\begin{align}
  \tstop\OmK \ll h_\mathrm{g}. 
\end{align} 
For thin PPDs with $h_\mathrm{g}\sim 0.05$, this is satisfied 
for $\tstop\OmK \ll O(10^{-2})$.

### 2016-08-24 summary.tex
general edits

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} = - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla \ln{K}
  + \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  s = \ln{\left[K^\frac{1}{\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas with minimal modification. In fact, no
modification is needed  
for strictly isothermal or polytropic gas ($c_s$ or $K$ being
constant, respectively) perfectly-coupled to dust. The usual 
adiabatic energy equation is effectively a dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
can simply add corresponding source terms in energy equation. In the
large but finite coupling case, this source term is analogous to
radiative diffusion \citep{price15}, which is also standard in many
hydrodynamics codes.

>>>>

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  s = \ln{\left[K^\frac{1}{\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one can set $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas with minimal modification. In fact, no
modification is needed  
for strictly isothermal or polytropic gas perfectly-coupled to dust:  
the standard adiabatic energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in energy equation. 
This source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15}, which is also
standard in many hydrodynamics codes.
